---
title: "Assignment week 5"
author: "Carla Ebner"
date: "2025-03-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(rtracklayer)
  library(epiwraps)
  library(GenomicRanges)
  library(R.utils)
})
ah <- AnnotationHub()
ensdb <- ah[["AH95713"]] # Drosophila EnsDb
```


# 1. Identify bivalent domains (H3K27me3 + H3K4me3) from (mouse or human) embryonic stem cells (ESC)
### First we download and unzip the files that we found on ENCODE


```{R}
download.file("https://www.encodeproject.org/files/ENCFF423ZVR/@@download/ENCFF423ZVR.bed.gz", "H3K27me3.bed.gz")
gunzip("H3K27me3.bed.gz", remove=TRUE)
```

```{R}
download.file("https://www.encodeproject.org/files/ENCFF589MNA/@@download/ENCFF589MNA.bed.gz", "H3K4me3.bed.gz")
gunzip("H3K4me3.bed.gz", remove=TRUE)
```
### Then we load all the peaks

```{R}
H3K27me3_peaks <- import("H3K27me3.bed", format="narrowPeak")
H3K4me3_peaks <- import("H3K4me3.bed", format="narrowPeak")
```

```{R}
unique(seqnames(H3K27me3_peaks))
unique(seqnames(H3K4me3_peaks))
```

### Now we want to identify the bivalent domains, so regions that are marked by H3K27me3 and H3K4me3 

```{R}
bivalent_H3K27me3_H3K4me3 <- findOverlaps(H3K27me3_peaks, H3K4me3_peaks, type="within")
bivalent_H3K27me3_H3K4me3
```


```{R}
bivalent_domains <- suppressWarnings(
  pintersect(H3K27me3_peaks[from(bivalent_H3K27me3_H3K4me3)], 
              H3K4me3_peaks[to(bivalent_H3K27me3_H3K4me3)])
)
bivalent_domains
```
### We found 8149 bivalent domains

# 2. What happens to those regions upon differentiation?
### We choose erythroblasts as differentiated cell type and download the files for H3K27me3 and H3K4me3 

```{R}
download.file("https://www.encodeproject.org/files/ENCFF827PLH/@@download/ENCFF827PLH.bed.gz", "H3K27me3_erythroblasts.bed.gz")
gunzip("H3K27me3_erythroblasts.bed.gz", remove=TRUE)
```

```{R}
download.file("https://www.encodeproject.org/files/ENCFF851JRN/@@download/ENCFF851JRN.bed.gz", "H3K4me3_erythroblasts.bed.gz")
gunzip("H3K4me3_erythroblasts.bed.gz", remove=TRUE)
```

### Then we load the peaks

```{R}
H3K27me3_erythroblasts_peaks <- import("H3K27me3_erythroblasts.bed", format="narrowPeak")
H3K4me3_erythroblasts_peaks <- import("H3K4me3_erythroblasts.bed", format="narrowPeak")
```

### Now we want to find the overlapping regions of this differentiated cell type (erythroblasts) with the mESC bivalent domains
### Overlaps with H3K27me3 erythroblasts region: 2829 overlaps

```{R}
H3K27me3_overlaps <- findOverlaps(bivalent_domains, H3K27me3_erythroblasts_peaks, type="within")
H3K27me3_overlaps
length(H3K27me3_overlaps)
```

### Overlaps with H3K4me3 erythroblasts region: 2260 overlaps

```{R}
H3K4me3_overlaps <- suppressWarnings(findOverlaps(bivalent_domains, H3K4me3_erythroblasts_peaks, type="within"))
H3K4me3_overlaps
length(H3K4me3_overlaps)
```

### Overlaps with both marks: 96 overlaps. Here we create the overlap of the H3K27me3 erythroblasts peaks and the H3K4me3 erythroblasts peaks first

```{R}
both <- findOverlaps(H3K27me3_erythroblasts_peaks, H3K4me3_erythroblasts_peaks, type="within")
both
```

```{R}
both_domains <- pintersect(H3K27me3_erythroblasts_peaks[from(both)], 
              H3K4me3_erythroblasts_peaks[to(both)])
both_domains
```


```{R}
both_overlaps <- findOverlaps(both_domains, bivalent_domains, type="within")
both_overlaps
length(both_overlaps)
```
### Now we want to show the results. We see that the biggest part of the bivalent domains, doesn't stay bivalent after differentiation

```{R}
cat("The number of mESC bivalent domains overlapping with H3K27me3 in erythroblasts is:", length(H3K27me3_overlaps), "\n") 

cat("The number of mESC bivalent domains overlapping with H3K4me3 in erythroblasts is:", length(H3K4me3_overlaps), "\n") 

cat("The number of mESC bivalent domains overlapping with both H3K27me3 and H3K4me3 in erythroblasts is:", length(both_overlaps), "\n") 
```


